Kernel Phase and Coronagraphy with Automatic Differentiation

### 2020-11-18 ms.tex

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
% optical imaging with aberrations
The accumulation of aberrations along the optical path in a telescope produces distortions and speckles in the resulting images, limiting the performance of cameras at high angular resolution. It is important to achieve the highest possible sensitivity to faint sources, using both hardware and data analysis software. While analytic methods are efficient, real systems are better-modelled numerically, but numerical models of complicated optical systems with many parameters can be hard to understand, optimize and apply.
% autodiff can make this practical for arbitrary optical systems
Automatic differentiation or `backpropagation' software developed for machine learning applications now makes calculating derivatives with respect to aberrations in arbitrary planes straightforward for any optical system.
% self cal - closure phase and kernel phase
We apply this powerful new tool to the problem of high-angular-resolution astronomical imaging. Self-calibrating observables such as the `closure phase' or `bispectrum' have been widely used in optical and radio astronomy to mitigate optical aberrations and achieve precise astrometry. Kernel phases are a generalization of closure phases valid in the limit of small phase errors.
% we demonstrate a simple example of the kernel of a monochromatic coronagraph
Using automatic differentiation, we reproduce existing kernel phase theory within this framework and demonstrate an extension to the case of a Lyot coronagraph, which is found to have self-calibrating combinations of speckles which are resistant to phase noise, but only in the very high wavefront quality regime. We apply reanalyze Palomar adaptive optics observations of the binary $\alpha$~Ophiuchi, finding consistency between the new pipeline and the existing standard.
%phase design
We present a Python package \textsc{morphine} that incorporates these ideas, with an interface similar to the popular package \textsc{poppy}, for optical simulation with automatic differentiation. These methods may be useful for designing improved astronomical optical systems by gradient descent.
\href{https://github.com/benjaminpope/morphine}{\color{linkcolor}\faGithub} % shamelessly borrowed from Luger!

@@

\section{Introduction} 
\label{sec:intro}
% optical imaging with aberrations
Many questions in astronomy and other sciences can only be answered with diffraction-limited high resolution imaging. The highest resolutions are typically achieved with the method of interferometry, in which waves detected at multiple receivers are combined physically or in post-processing to obtain the Fourier transform of the source intensity distribution \citep{vc34,zernike38}. Even in the case of a single telescope or camera, it is often nevertheless helpful to think of them as an interferometer composed of many sub-apertures which combine their signals directly onto a focal plane \citep[a `Fizeau interferometer':][]{fizeau1868}. In all of these cases, unknown path delays toward each receiver or subaperture are the dominant source of noise, causing `tip-tilts' of the final image position, distortions of the point spread function (PSF) from low-spatial-frequency aberrations, and clouds of `speckles' from high-spatial-frequency aberrations.

% self cal with closure phase
In order to correct for these aberrations, it is possible to use active hardware such as adaptive optics or delay lines; calibrating these errors with comparison to a reference star; or self-calibration, using the physics of the noise process to correct for it in post-processing.
In the recovery of object phase information, among the most longstanding techniques is that of self-calibration with `closure phases' \citep[introduced in the context of radio astronomy by][]{jennison58}, in which phases are summed around three interferometric baselines which form a closing triangle. The phase error terms local to each subaperture cancel, so that three low-signal-to-noise (SNR) baseline phases deliver one high-SNR observable. In a non-redundant array (one in which no baseline vector is repeated between different pairs of subapertures) of sufficient size, a large number of closure phases can be obtained which anchor image reconstruction with great precision \citep{chael18}. `Closure amplitudes' which are resistant to fluctuations in input amplitude or gain can also be obtained by a similar construction using four telescopes \citep{twiss60,blackburn20}. The closure phase is the argument of a quantity called the triple product or `bispectrum', and it has recently been shown from the perspective of invariant theory in algebraic geometry that for a wide class of problems limited by phase noise, knowledge of the mean of a signal, its power spectrum, and its bispectrum are necessary and sufficient for an optimal signal reconstruction \citep{bandeira17}.

@@

% introduce kernel phase idea
The kernel phase method is a way of extending closure phase from simple nonredundant arrays of telescopes to the densely-filled pupils of real telescopes. If we describe the propagation of phase and amplitude noise in terms of a matrix \citep{lannes1991}, we can obtain powerful generalizations to closure phases and amplitudes. We will write describe the effect of aberrations on baseline phases as

@@

\subsection{Automatic Differentiation}
% autodiff/backpropagation - history & citations
The practical use of neural networks in machine learning applications is dependent on the efficient calculation of analytic gradients of often very complicated composite functions, for example the matrix operations composed with nonlinear activation functions that are seen in neural networks \citep{lecun15}. This problem is usually referred to as algorithmic differentiation, automatic differentiation or `autodiff', and is solved simply by the chain rule.
Autodiff is available in many implementations, such as the Python packages \textsc{TensorFlow} \citep{tensorflow2015}, \textsc{theano} \citep{theano}, \textsc{PyTorch} \citep{pytorch}, \textsc{autograd} \citep{autograd}, \textsc{jax} \citep{jax}, and many packages in the Julia language \citep{julia}. The reverse-mode autodiff or `backpropagation' algorithm \citep{linnainmaa1970,lecun1988theoretical} has made this practical for neural networks, but it is typically most-useful in cases where the output dimensionality is much smaller than the input, such as neural networks; forwards-mode autodiff, on the other hand, is usually better in the case where the output is of higher dimension than the input. For optics problems, both of these can be the case - for example, for optimization, backpropagation is usually the best approach, but for kernel phase analysis, forwards mode is better suited.

@@

For our forwards model, we adapt the popular physical optics library \textsc{poppy} \citep{poppy}, with the goal of compatibility with existing simulations built on \textsc{poppy} such as \textsc{WebbPSF} \citep{webbpsf}. We use the matrix Fourier transform mode to avoid using the FFT \citep{soummer07}, to more easily generate arbitrary image sampling and for consistency with the kernel phase code \textsc{xara} \citep{martinache20}. We have adapted the low-level features of \textsc{poppy} to use the Google autodiff library \textsc{jax} \citep{jax} in place of \textsc{NumPy}, and to distinguish it from the original version we call this new \textsc{poppy} derivatives library \textsc{morphine}. %The \textsc{NumPy}-like API of \textsc{jax} is chosen so that astronomers can painlessly integrate \textsc{morphine} with their existing code, but we note that a \textsc{DeepOptics}-like approach using \textsc{TensorFlow} may be preferable in the longer term for integration with neural networks.

@@

\subsection{Simple Diffraction}
\label{sec:simple}

First we want to reproduce the basic kernel phase calculations in this new model. We propagate monochromatic 2.0\,$\mu$m light through a 2.0\,m diameter circular pupil onto a detector with a 20\,mas/pixel scale and 4\,arcsec field of view. The pupil and $u,v$ are each calculated on a 64x64 grid. We evaluate the Jacobian of the $u,v$ phases with respect to the input pupil phases using forwards-mode autodiff and then evaluate this Jacobian at zero phase to get our $\mathbf{A_\phi}$ matrix. On a laptop computer this takes a few minutes processing time. Memory usage is the main overhead: in practice on a laptop with 16GB RAM, we are limited to grids no larger than 128x128. Demand on computational resources presents a major limitation to widespread application; in future this may be overcome by more efficient refactoring of the code, or simply by doing calculations on clusters with very large RAM allocations.

@@

In Figure~\ref{kernel_jacobian} we display the results of our Jacobian calculation for a set of highlighted pupil samples and how they influence the phases in the $u,v$ plane.

@@

% \footnote{The OTF is the autocorrelation of the pupil, and is in this case also a circle. Its support is calculated by calculating a $u,v$ forwards model with zero phase offset and making an amplitude-threshold mask.}
\begin{figure}
\plotone{jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for monochromatic 2$\mu$m diffraction from a 2.0\,m diameter circular pupil. In each pair of images, on the left is the pupil, with a red dot highlighting a selected pixel; on the right is a map of the derivative in the $u,v$ plane with respect to the phase at this point in the pupil. We see the same maps as in Figure~1 of \citet{martinache10}. \label{kernel_jacobian}}
\end{figure}

\begin{figure}
    \plotone{kerphi_singular.pdf}
     \caption{Log-scale plot of the singular values of the phase transfer matrix plotted against index ordered by decreasing singular value for simple monochromatic diffraction (blue), 20\% bandwidth (green). Singular values are normalized to the first singular value. In both cases, there is a sharp cutoff separating significant singular values from a subset within machine precision of zero. The vectors corresponding to these approximately zero singular values span the null space of the phase transfer operator, and correspond to kernel phases. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_uv.ipynb}{\color{linkcolor}\faGithub}}
    \label{fig:svd}
\end{figure}

\begin{figure}
\plotone{nonsingular_modes_mono.pdf}
\caption{Pairs of nonsingular vectors in the pupil and $u,v$ planes from the SVD of the Jacobian in ~\ref{kernel_jacobian}. Some of these orthonormal basis vectors resemble Zernike modes, and all show some ringing structure at the edges of the pupil and OTF support.\href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_uv.ipynb}{\color{linkcolor}\faGithub}
\label{nonsingular_modes}}
\end{figure}

\begin{figure}
\plotone{kernel_modes_mono.pdf}
\caption{Some kernel phase maps - singular vectors in the $u,v$ plane from the SVD of the Jacobian in ~\ref{kernel_jacobian}. There is little apparent structure in these.\href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_uv.ipynb}{\color{linkcolor}\faGithub}
\label{kernel_modes}}
\end{figure}

@@

\subsection{$\alpha$~Ophiuchi}
\label{sec:palomar}

We now apply the new kernel phase formalism to real data: the A-star $\alpha$~Ophiuchi. It was observed with the Palomar~200-Inch telescope, using the extreme adaptive optics system PALM-3K and the PHARO camera. These data were first analyzed by \citet{palomar}, but \citet{martinache20} showed that there was an error in the pupil model used for the original analysis: the pupil was in fact rotated 2 degrees. Using a more accurate pupil model and the updated `xara' kernel phase code, \citet{martinache20} obtained astrometry of separation $\rho = 123.5 \pm 2.9$~mas, position angle $\theta = 86.5 \pm 0.2$~degrees, and contrast $c = 25.1 \pm 1.1$. We therefore seek to recreate this new analysis using autodiff.

@@

\begin{figure}
\plotone{pharo_jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for the PHARO pupil in $K$-band. In each pair of images, on the left is the PHARO pupil, with a red dot highlighting a pupil sample; on the right is a map of the derivative in the $u,v$ plane with respect to phase at that pupil sample. We see the same maps as in Figure~1 of \citet{martinache10}. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_pharo_xara.ipynb}{\color{linkcolor}\faGithub}\label{pharo_jacobian}}
\end{figure}

We then extract kernel phases from both $\alpha$~Oph and a point source calibrator $\epsilon$~Her using a version of \textsc{xara} modified for compatibility with \textsc{morphine}. Following \citet{martinache20}, we choose as our observables the median kernel phases across all 100 exposures for each, and the standard error of the mean as our base uncertainty on each. To calibrate the kernel phases, we simply subtract those of the calibrator from $\alpha$~Oph and add their uncertainties in quadrature. To see the effect of the known optical `ghost' (introduced by an unwanted reflection from the neutral density filter used to observe bright stars), we calculate kernel phase `colinearity' maps as in \citet{martinache20}, shown in Figure~\ref{colinearity}. In these maps we can see that the filter ghost shows up strongly as a false binary in the uncalibrated kernel phase maps, but calibrated kernel phases easily remove this and reveal the tight $\alpha$~Oph binary.

In \citet{martinache20} a large error term was added in quadrature in order to account for additional noise in the data unexplained by the diversity over exposures and uncalibrated by the kernel phase model. Using nonlinear least squares to fit the calibrated data, we also find that the best-fitting binary model has a high $\chi^2$. We therefore follow \citet{martinache20} and add a dimensionless kernel phase quantity of 0.0184 to in quadrature to the uncertainties to make the reduced $\chi^2 = 1.0$.

\begin{figure}
\plotone{colinearity_alphaoph.pdf}
\caption{Kernel phase colinearity maps showing the normalized dot product of a binary model (pale is a better match) to the kernel phase signals from uncalibrated $\alpha$~Oph data (left), the calibrator $\eps$~Her (middle), and calibrated data (right). They closely resemble the maps in Figure~8 of \citet{martinache20}. We can see that in uncalibrated data the filter ghost is clearly visible, but in calibrated data we extract the $\alpha$~Oph binary with no effect from the ghost. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/pharo_test.ipynb}{\color{linkcolor}\faGithub} \label{colinearity}}
\end{figure}

@@

% In order to investigate this discrepancy, we generated a grid of forwards models. Using the same code used to produce the kernel phase autodiff model, we generated input wavefronts with 20 Zernike modes of aberrations, each with random amplitudes normally-distributed with 15\,nm standard deviation, for a total peak-to-peak amplitude of 206\,nm. These were then shifted and added to simulate a range of binaries at different configurations and contrasts, and both the \citet{martinache20} pipeline and the autodiff pipeline were used to retrieve their parameters. The results are shown in Figure~\ref{method_comparison_ensemble}. The pipelines achieve broadly comparable performance on these simulations, with the \citet{martinache20} method being slightly better at closer separations and higher contrasts and autodiff performing marginally better at larger separations.

In order to investigate the effect on both pipelines of wavefront errors leaking in, we generated a cube of 1000 PSFs with 20 Zernike modes of aberrations, this time with 20\,nm standard deviation, for a total peak-to-peak wavefront amplitude of $295\pm 100$\,nm. We then retrieved the system parameters with \texttt{emcee} for the cube
%, taking as our observables the median kernel phases and the standard error of the mean over the ensemble (posterior in Figure~\ref{comparison_posterior_sim}); and also 
frame-by-frame using weighted least squares (displayed in Figure~\ref{comparison_posterior_sim_cube}). For a simulated binary with fiducial $\rho = 125$\,mas, $\theta=86^\circ$, and $c = 25$, there is no sign of the large systematic difference between the methods as seen in real data. Both methods, whether frame by frame or on average, retrieve the true parameters with similar accuracy and precision. %Whatever systematic this is may be treated differently by these two pipelines, and may need to be a subject for future work.

% \begin{figure}
% \plotone{method_comparison_ensemble.pdf}
% \caption{Comparison of \citet[][blue stars]{martinache20} and autodiff (orange dots) kernel phase pipelines on simulated data, as a function of input contrast (left) and separation (right). When inspected, the samples with worse separation error after both pipelines are those at higher contrasts. Uncertainties are omitted for clarity, and because they are poorly defined in this context. \label{method_comparison_ensemble}}
% \end{figure}

% \begin{figure}
% \plotone{comparison_posterior_sim.pdf}
% \caption{Comparison of \citet[][blue linetype]{martinache20} and autodiff (black) kernel phase pipelines on simulated data with true parameters $\rho = 125$~mas, $\theta = 86^\circ$, $c=25$. We generate a cube of 1000 mock datasets using a 64-pix pupil model and Zernike aberrations as described in Section~\ref{sec:palomar}, and the observables used are the median and standard error of the mean of each kernel phase across this ensemble. Both datasets retrieve the true parameters very precisely, though with small systematic offsets, which are slightly larger for the autodiff method than for classic kernel phases. \label{comparison_posterior_sim}}
% \end{figure}

@@

In order to establish the accuracy of both pipelines, we conducted an injection test. We took the first frame of the $\eps$~Her datacube as our `data' and use a subpixel Fourier shift to create a mock binary at a position angle of $90^\circ$, contrast 25, and a range of separations from 90-300~mas. We then used least squares to fit these data using both models (initialized at the true values), determined in each case an error to add in quadrature to make reduced $\chi^2=1.0$, and re-ran the least squares fit with new data, saving the best-fit parameters and uncertainties taken from the square root of the diagonal elements of the inverse covariance matrix. The results are shown in Figure~\ref{injection_recovery}. Both methods perform comparably well, achieving results reasonably close to the injected values. Systematics in all three parameters, of amplitude comparable to the $1\sigma$ statistical uncertainties, vary smoothly and similarly as a function of input separation. The autodiff pipeline performs somewhat better at smaller separations in all three parameters. This indicates that both kernel phase pipelines do a good overall job at binary imaging with PHARO, but that as noted by \citet{martinache20} not all sources of uncertainty are well-understood or calibrated.

@@

% In numerical experiments, subpixel mismatch in Fourier transform coordinates or even Fourier centering convention could cause mismatches between simulation parameters and retrievals at around the scale of the discrepancy between the two methods.  These results are in line with the \citet{martinache_habilitation} conclusion that inexact Fourier sampling contributes significant systematic noise to kernel phases. The final version used for the autodiff pipeline uses the exact optical model used to generate the data, and a differentiable version of the exact $u,v$ transform implementation used in \textsc{xara}; we do not believe this to be the main cause of the small discrepancy between autodiff and conventional parameters, but it does illustrate that model mis-specification may be a contributing factor.

\subsection{Coronagraphy}
\label{sec:coronagraph}

% coronagraphs, speckles etc
It is desirable to find a priori self-calibration schemes for high-angular-resolution, high-contrast imaging systems more generally, including not merely nulling interferometers but optically complex coronagraphs intended to suppress light in a region around a target star. One particularly interesting extension of the kernel phase idea is the `kernel nuller' concept \citep{martinache18}, in which the aberration transfer matrix and kernel operator idea is applied to the output of a nulling interferometer.

@@

Here we will apply the formalism demonstrated above for simple diffraction to a new case: coronagraphy. We will find that the exact same approach generates an analogous modal decomposition and kernel observables.

@@

Calculating the SVD of this Jacobian (again restricted to the support of the primary pupil) we see a decline to very small singular values, albeit not as steep as earlier for the simple imaging case. The ordered singular value curve is displayed in Figure~\ref{fig:svd_coronagraph}. The nonsingular modes form a basis, like the Zernike basis, that maps between aberration patterns in the pupil and the speckle modes they generate. Some pairs of nonsingular modes in the pupil and image plane are shown in Figure~\ref{nonsingular_corona}

\begin{figure}
\plotone{speckle_jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for a simple Lyot coronagraph. In each pair of images, on the left is the speckle pattern, with a red dot highlighting the selected pixel; on the right is a map of the corresponding pixel's derivative with respect to phase over the pupil. Speckles far from the central dark hole correspond simply to sinusoids across the pupil, while speckles near the inner working angle correspond to more complicated patterns.  \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_coronagraph.ipynb}{\color{linkcolor}\faGithub}
\label{speckle_jacobian}}
\end{figure}

>>>>

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
% optical imaging with aberrations
The accumulation of aberrations along the optical path in a telescope produces distortions and speckles in the resulting images, limiting the performance of cameras at high angular resolution. It is important to achieve the highest possible sensitivity to faint sources, using both hardware and data analysis software. While analytic methods are efficient, real systems are better modelled numerically, but numerical models of complicated optical systems with many parameters can be hard to understand, optimize and apply.
% autodiff can make this practical for arbitrary optical systems
Automatic differentiation or `backpropagation' software developed for machine learning applications now makes calculating derivatives with respect to aberrations in arbitrary planes straightforward for any optical system.
% self cal - closure phase and kernel phase
We apply this powerful new tool to the problem of high-angular-resolution astronomical imaging. Self-calibrating observables such as the `closure phase' or `bispectrum' have been widely used in optical and radio astronomy to mitigate optical aberrations and achieve high fidelity imagery. Kernel phases are a generalization of closure phases valid in the limit of small phase errors.
% we demonstrate a simple example of the kernel of a monochromatic coronagraph
Using automatic differentiation, we reproduce existing kernel phase theory within this framework and demonstrate an extension to the case of a Lyot coronagraph, which is found to have self-calibrating combinations of speckles which are resistant to phase noise, but only in the very high wavefront quality regime. 
As an illustrative example, we reanalyze Palomar adaptive optics observations of the binary $\alpha$~Ophiuchi, finding consistency between the new pipeline and the existing standard.
%phase design
We present a new Python package \textsc{morphine} that incorporates these ideas, with an interface similar to the popular package \textsc{poppy}, for optical simulation with automatic differentiation. These methods may be useful for designing improved astronomical optical systems by gradient descent.
\href{https://github.com/benjaminpope/morphine}{\color{linkcolor}\faGithub} % shamelessly borrowed from Luger!

@@

\section{Introduction} 
\label{sec:intro}
% optical imaging with aberrations
Many questions in astronomy and other sciences can only be answered with diffraction-limited high resolution imaging. The highest resolutions are typically achieved with the method of interferometry, in which waves detected at multiple receivers are combined physically or in post-processing to obtain the Fourier transform of the source intensity distribution \citep{vc34,zernike38}. Even in the case of a single telescope or camera, it is often nevertheless helpful to think of them as an interferometer composed of many sub-apertures which combine their signals directly onto a focal plane \citep[a `Fizeau interferometer':][]{fizeau1868}. In all of these cases, unknown path delays through each receiver or subaperture are the dominant source of aberration with low spatial frequency structure causing issues such as varying `tip-tilt' of the final image position, while higher order distortions yield clouds of `speckles'.

% self cal with closure phase
Strategies to correct for these aberrations include the use of active systems such as adaptive optics or delay lines; calibrating these errors with comparison to a reference star; or self-calibration, using the physics of the noise process to correct for it in post-processing.
In the recovery of object phase information, among the most longstanding techniques is that of self-calibration with `closure phases' \citep[introduced in the context of radio astronomy by][]{jennison58}, in which phases are summed around three interferometric baselines which form a closing triangle. The phase error terms local to each subaperture cancel, so that three low-signal-to-noise (SNR) baseline phases deliver one high-SNR observable. In a non-redundant array (one in which no baseline vector is repeated between different pairs of subapertures) of sufficient size, a large number of closure phases can be obtained which anchor image reconstruction with great precision \citep{chael18}. `Closure amplitudes' which are resistant to fluctuations in input amplitude or gain can also be obtained by a similar construction using four telescopes \citep{twiss60,blackburn20}. The closure phase is the argument of a quantity called the triple product or `bispectrum', and it has recently been shown from the perspective of invariant theory in algebraic geometry that for a wide class of problems limited by phase noise, knowledge of the mean of a signal, its power spectrum, and its bispectrum are necessary and sufficient for an optimal signal reconstruction \citep{bandeira17}.

@@

% introduce kernel phase idea
The kernel phase method is a way of extending closure phase from simple nonredundant arrays of telescopes to the densely-filled pupils of real telescopes. If we describe the propagation of phase and amplitude noise in terms of a matrix \citep{lannes1991}, we can obtain powerful generalizations to closure phases and amplitudes. We will describe the effect of aberrations on baseline phases as

@@

\subsection{Automatic Differentiation}
% autodiff/backpropagation - history & citations
The practical use of neural networks in machine learning applications is dependent on the efficient calculation of analytic gradients of often very complicated composite functions, for example the matrix operations composed with nonlinear activation functions that are seen in neural networks \citep{lecun15}. This problem is usually referred to as algorithmic differentiation, automatic differentiation or `autodiff', and is solved simply by the chain rule.
Autodiff is available in many implementations, such as the Python packages \textsc{TensorFlow} \citep{tensorflow2015}, \textsc{theano} \citep{theano}, \textsc{PyTorch} \citep{pytorch}, \textsc{autograd} \citep{autograd}, \textsc{jax} \citep{jax}, and many packages in the Julia language \citep{julia}. The reverse-mode autodiff or `backpropagation' algorithm \citep{linnainmaa1970,lecun1988theoretical} has made this practical for neural networks, but it is typically most-useful in cases where the output dimensionality is much smaller than the input, as for most neural networks. Forwards-mode autodiff, on the other hand, is usually better in the case where the output is of higher dimension than the input. For optics problems, both cases can be found -- for example, for optimization, backpropagation is usually the best approach, while for kernel phase analysis, forwards mode is better suited.

@@

To forwards-model systems, we adapt the popular physical optics library \textsc{poppy} \citep{poppy}, with the goal of compatibility with existing simulations built on \textsc{poppy} such as \textsc{WebbPSF} \citep{webbpsf}. We use the matrix Fourier transform mode and avoid FFTs \citep{soummer07} to more easily generate arbitrary image sampling and for consistency with the kernel phase code \textsc{xara} \citep{martinache20}. We have adapted the low-level features of \textsc{poppy} to use the Google autodiff library \textsc{jax} \citep{jax} in place of \textsc{NumPy}, and to distinguish it from the original version we call this new \textsc{poppy} derivatives library \textsc{morphine}. %The \textsc{NumPy}-like API of \textsc{jax} is chosen so that astronomers can painlessly integrate \textsc{morphine} with their existing code, but we note that a \textsc{DeepOptics}-like approach using \textsc{TensorFlow} may be preferable in the longer term for integration with neural networks.

@@

\subsection{Simple Diffraction Example}
\label{sec:simple}

First we want to ensure basic kernel phase calculations are reproduced in this new model. We propagate monochromatic 2.0\,$\mu$m light through a 2.0\,m diameter circular pupil that is imaged onto a detector with a 20\,mas/pixel scale and 4\,arcsec field of view. The pupil plane and the baseline $u,v$ plane are each calculated on a 64x64 grid. We evaluate the Jacobian of the $u,v$ phases with respect to the input pupil phases using forwards-mode autodiff and then evaluate this Jacobian at zero phase to get our $\mathbf{A_\phi}$ matrix. On a laptop computer this takes a few minutes processing time. Memory usage is the main overhead: with 16GB RAM, grid size is limited to about 128x128. Demand on computational resources presents a major limitation to widespread application; in future this may be overcome by more efficient refactoring of the code, or simply by doing calculations on clusters with very large RAM allocations.

@@

In Figure~\ref{kernel_jacobian} we display the results of our Jacobian calculation for a set of highlighted pupil samples and illustrating the influence of phase in the $u,v$ plane.

@@

% \footnote{The OTF is the autocorrelation of the pupil, and is in this case also a circle. Its support is calculated by calculating a $u,v$ forwards model with zero phase offset and making an amplitude-threshold mask.}
\begin{figure}
\plotone{jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for monochromatic 2$\mu$m diffraction from a 2.0\,m diameter circular pupil. In each pair of images, on the left is the pupil, with a red dot highlighting a selected pixel; on the right is a map of the derivative of the phase in the $u,v$ plane with respect to the phase at that point in the pupil. These Jacobian vectors are pairs of rows and columns of the $\mathbf{A}$ phase transfer matrix. We see the same maps as in Figure~1 of \citet{martinache10}. \label{kernel_jacobian}}
\end{figure}

\begin{figure}
    \plotone{kerphi_singular.pdf}
     \caption{Log-scale plot of the singular values of the phase transfer matrix plotted against index ordered by decreasing singular value for simple monochromatic diffraction (blue), 20\% bandwidth (green). Singular values are normalized to the first singular value. In both cases, there is a sharp cutoff separating significant singular values from a plateau at $\sim 5 \times10^{-8}$ (zero to within machine precision). We take the vectors corresponding to these singular values in the plateau region to be the effective null space of the phase transfer operator, and correspond to kernel phases. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_uv.ipynb}{\color{linkcolor}\faGithub}}
    \label{fig:svd}
\end{figure}

\begin{figure}
\plotone{nonsingular_modes_mono.pdf}
\caption{Pairs of nonsingular vectors in the pupil and $u,v$ planes from the SVD of the Jacobian in ~\ref{kernel_jacobian}. Some of these orthonormal basis vectors resemble Zernike modes, and all show some ringing structure at the edges of the pupil and OTF support. All eight images are dimensionless and the colour scale is arbitrary. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_uv.ipynb}{\color{linkcolor}\faGithub}
\label{nonsingular_modes}}
\end{figure}

\begin{figure}
\plotone{kernel_modes_mono.pdf}
\caption{Example kernel phase maps - singular vectors in the $u,v$ plane from the SVD of the Jacobian in ~\ref{kernel_jacobian}. These are rows of the $\mathbf{K}$ matrix. There is little apparent structure. All sixteen images are dimensionless and the colour scale is arbitrary. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_uv.ipynb}{\color{linkcolor}\faGithub}
\label{kernel_modes}}
\end{figure}

@@

\subsection{Stellar imaging data: $\alpha$ Ophiuchi}
\label{sec:palomar}

We now apply our new kernel phase formalism to real data: the A-type binary star system $\alpha$~Ophiuchi observed with the Palomar~200-Inch telescope using the extreme adaptive optics system PALM-3K and the PHARO camera. These data were first analyzed by \citet{palomar}, successfully recovering the binary and properties of the system. 
These were subsequently improved by \citet{martinache20} who corrected a 2 degree offset in the clocking angle of the real Palomar pupil. Also employing updated `xara' kernel phase code, \citet{martinache20} obtained astrometry of separation $\rho = 123.5 \pm 2.9$~mas, position angle $\theta = 86.5 \pm 0.2$~degrees, and contrast $c = 25.1 \pm 1.1$. The well-analyzed nature of these data make it something of a useful test-piece; we therefore seek to recreate this analysis using our new autodiff methods.

@@

\begin{figure}
\plotone{pharo_jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for the PHARO pupil in $K$-band. In each pair of images, on the left is the PHARO pupil, with a red dot highlighting a pupil sample; on the right is a map of the derivative in the $u,v$ plane with respect to phase at that pupil sample. These Jacobian vectors are pairs of rows and columns of the $\mathbf{A}$ phase transfer matrix. We see the same maps as in Figure~1 of \citet{martinache10}. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_pharo_xara.ipynb}{\color{linkcolor}\faGithub}\label{pharo_jacobian}}
\end{figure}

We then extract kernel phases from both $\alpha$~Oph and a point source calibrator $\epsilon$~Her using a version of \textsc{xara} modified for compatibility with \textsc{morphine}. Following \citet{martinache20}, we choose as our observables the median kernel phases across all 100 exposures for each, and the standard error of the mean as our base uncertainty on each. To calibrate the kernel phases, we simply subtract those of the calibrator from $\alpha$~Oph and add their uncertainties in quadrature. To see the effect of the known optical `ghost' (due to an unwanted reflection from the neutral density filter used to observe bright stars such as these), we calculate kernel phase `colinearity' maps as in \citet{martinache20}, shown in Figure~\ref{colinearity}. In these maps we can see that the filter ghost shows up strongly as a false binary in the uncalibrated kernel phase maps, but calibrated kernel phases easily remove this and reveal the tight $\alpha$~Oph binary.

In \citet{martinache20} a large error term was added in quadrature in order to account for additional noise in the data unexplained by the diversity over exposures and uncalibrated by the kernel phase model. Using nonlinear least squares to fit the calibrated data, we also find that the best-fitting binary model has a high $\chi^2$. We therefore follow \citet{martinache20} and add a dimensionless kernel phase quantity of 0.0184 in quadrature to the uncertainties to make the reduced $\chi^2 = 1.0$.

\begin{figure}
\plotone{colinearity_alphaoph.pdf}
\caption{Kernel phase colinearity maps showing the normalized dot product of a binary model (pale is a better match) to the kernel phase signals from uncalibrated $\alpha$~Oph data (left), the calibrator $\epsilon$~Her (middle), and calibrated data (right). They closely resemble the maps in Figure~8 of \citet{martinache20}. We can see that in uncalibrated data the filter ghost is clearly visible, but in calibrated data we extract the $\alpha$~Oph binary with no effect from the ghost. \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/pharo_test.ipynb}{\color{linkcolor}\faGithub} \label{colinearity}}
\end{figure}

@@

\subsection{Simulated binary star data}

In order to investigate the effect on both pipelines of wavefront errors leaking in, we generated a cube of 1000 PSFs at a wavelength of $2.145\,\mu$m with 20 Zernike modes of aberrations, this time with 20\,nm standard deviation, for a total peak-to-peak wavefront amplitude of $295\pm 100$\,nm. We then retrieved the system parameters with \texttt{emcee} for the cube
%, taking as our observables the median kernel phases and the standard error of the mean over the ensemble (posterior in Figure~\ref{comparison_posterior_sim}); and also 
frame-by-frame using weighted least squares (displayed in Figure~\ref{comparison_posterior_sim_cube}). For a simulated binary with fiducial $\rho = 125$\,mas, $\theta=86^\circ$, and $c = 25$, there is no sign of the large systematic difference between the methods as seen in real data. Both methods, whether frame by frame or on average, retrieve the true parameters with similar accuracy and precision. %Whatever systematic this is may be treated differently by these two pipelines, and may need to be a subject for future work.

@@

In order to establish the accuracy of both pipelines, we conducted an injection test. We took the first frame of the $\epsilon$~Her datacube as our `data' and use a subpixel Fourier shift to create a mock binary at a position angle of $90^\circ$, contrast 25, and a range of separations from 90-300~mas. We then used least squares to fit these data using both models (initialized at the true values), determined in each case an error to add in quadrature to make reduced $\chi^2=1.0$, and re-ran the least squares fit with new data, saving the best-fit parameters and uncertainties taken from the square root of the diagonal elements of the inverse covariance matrix. The results are shown in Figure~\ref{injection_recovery}. Both methods perform comparably well, achieving results reasonably close to the injected values. Systematics in all three parameters, of amplitude comparable to the $1\sigma$ statistical uncertainties, vary smoothly and similarly as a function of input separation. The autodiff pipeline performs somewhat better at smaller separations in all three parameters. This indicates that both kernel phase pipelines do a good overall job at binary imaging with PHARO, but that as noted by \citet{martinache20} not all sources of uncertainty are well-understood or calibrated.

@@

\subsection{Complex optical systems: nullers and coronagraphs}
\label{sec:coronagraph}

% coronagraphs, speckles etc
It is desirable to find a priori self-calibration schemes for high-angular-resolution, high-contrast imaging systems more generally such as nulling interferometers and more optically complex coronagraphs intended to suppress light in a region around a target star. One particularly interesting extension of the kernel phase idea is the `kernel nuller' concept \citep{martinache18}, in which the aberration transfer matrix and kernel operator idea is applied to the output of a nulling interferometer.

@@

Here we will apply the formalism demonstrated above for simple diffraction to a new case: coronagraphy. We will find that the identical approach generates an analogous modal decomposition and kernel observables.

@@

Calculating the SVD of this Jacobian (again restricted to the support of the primary pupil) we see a decline to very small singular values. The ordered singular value curve is displayed in Figure~\ref{fig:svd_coronagraph}. The nonsingular modes form a basis, like the Zernike basis, that maps between aberration patterns in the pupil and the speckle modes they generate. Some pairs of nonsingular modes in the pupil and image plane are shown in Figure~\ref{nonsingular_corona}

\begin{figure}
\plotone{speckle_jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for a simple Lyot coronagraph. In each pair of images, on the left is the PSF, with a red dot highlighting the selected pixel; on the right is a map of the corresponding pixel's derivative with respect to phase over the pupil, the pattern of pupil phases that linearly affect that speckle. Speckles far from the central dark hole are generated by sinusoids across the pupil, while speckles near the inner working angle correspond to more complicated patterns.  \href{https://github.com/benjaminpope/morphine/blob/stable/notebooks/morphine_coronagraph.ipynb}{\color{linkcolor}\faGithub}
\label{speckle_jacobian}}
\end{figure}
